---
title: Fur seal odour encodes colony membership, mother-offspring similarity, relatedness
  and genetic quality
author: Stoffel, M.A., Caspers, B.A., Forcada, J., Giannakara, A., Baier, M.C., Eberhart-Phillips,
  L.J. , Müller, C. & Hoffman, J.I.
output:
  html_document:
    highlight: pygments
  pdf_document: default
---
!!!This file is under developement!!! and will soon contain all major analyses.

This is a script providing the code for all major analyses in our paper, follwing
the order of analyses ni the results section. Some large
functions are outsourced into a package that can be loaded into R with the 
follwing code.

```{r}
library(devtools)
# install_github()
```

***
### Loading data, standardisation and transformation

Loading the raw scent data (aligned by algorithm) and a factor data frame containing identities
for colony membership (colony), mother-offspring pairs (family) and mothers
and pups, respectively (age)

```{r}
scent_raw <- as.data.frame(t(read.csv("scent_raw.csv", row.names = 1)))
factors <- read.csv("factors.csv",row.names=1) 
names(factors)  <- c("colony", "family")
factors$age <- c(rep(1,41), rep(2,41))
```

Standardising observations by total, such that within every observation compounds
add up to 100 % (Thus averaging out absolute concentration differences between samples)

```{r}
scent_stand <- as.data.frame(t(apply(scent_raw, 1, function(x) (x/sum(x)) * 100)))
```

Log(x+1) transformation of the standrardised scent data.
```{r}
scent <- log(scent_stand + 1)
```

The scent matrix contains 82 observations and 213 compounds (retention times are column names) in total
```{r}
dim(scent)
head(scent[1:6])
head(factors)
```

***
### Differences in olfactory fingerprints between the two colonies

```{r, results="hide", message=FALSE}
library(vegan)
library(MASS)
```

Plot nMDS
```{r, fig.height=6, fig.width=6, warning = FALSE}
scent_mds <- isoMDS(vegdist(scent))
ordiplot(scent_mds, type = "t", ylab = "", xlab = "",axes=FALSE, frame.plot=TRUE)
```

Analysis of Similarities (ANOSIM) tests for group differences based on a Bray-curtis similarity matrix.

**Dissimilarity between the two colonies.**
```{r, cache=TRUE}
anosim(dat = scent, grouping = factors$colony, 
       distance = "bray", permutations = 1000)
```

**Dissimilarity between mothers from the two colonies.**
```{r, cache=TRUE}
anosim(dat = scent[factors$age == 1, ], grouping = factors$colony, 
       distance = "bray", permutations = 1000)
```

**Dissimilarity between pups from the two colonies.**
```{r, cache=TRUE}
anosim(dat = scent[factors$age == 2, ], grouping = factors$colony, 
       distance = "bray", permutations = 1000)
```

*Genetic differentiation of the two colonies was tested in Arlequin (Fst) and
Structure (Bayesian structure analysis)*  

***
### Mother offspring similarity

**Overall**
```{r, cache=TRUE}
anosim(dat = scent, grouping = factors$family, 
       distance = "bray", permutations = 1000)
```

**Within colony 1 (Special study beach)**
```{r, cache=TRUE}

anosim(dat = scent[factors$colony == 1, ], 
       grouping = factors[factors$colony == 1, ]$family, 
       distance = "bray", permutations = 1000)
```
**Within colony 2 (Freshwater beach)**
```{r, cache=TRUE}
anosim(dat = scent[factors$colony == 2, ], 
       grouping = factors[factors$colony == 2, ]$family, 
       distance = "bray", permutations = 1000)
```
  

***
### Olfactory similarity vs. geographic distance on special study beach
  
For the special study beach, exact data of the geographical positions is available.    
```{r}
coord  <- read.csv("coordinates_beach1.csv", row.names=1) 
```

Converting coordinates to pairwise euclidian distance matrix
```{r}
dist_mat <- as.matrix(dist(coord, method = "euclidian"))
```

Constructing bray curtis similarity matrix of all individuals from beach 1
```{r}
scent_bc <- as.matrix(vegdist(as.matrix(scent[factors$colony == 1, ])),
                      method = "bray")
```

**Geographic distance vs. olfactory similarity in mothers**
```{r}
geo_mum <- dist_mat[1:20, 1:20]
scent_mum <- scent_bc[1:20, 1:20]
vegan::mantel(geo_mum, scent_mum, method = "spearman")
```

**Geographic distance vs. olfactory similarity in pups**
```{r}
geo_pup <- dist_mat[21:40, 21:40]
scent_pup <- scent_bc[21:40, 21:40]
vegan::mantel(geo_pup, scent_pup, method = "spearman")
```

***
## Relatedness and overall olfactory similarity

Load pairwise relatedness based on 41 markers
```{r}
relatedness <- as.matrix(read.csv("relatedness.csv",row.names=1))
```

Pairwise bray curtis similarity of all individuals
```{r}
scent_bc <- 1-(as.matrix(vegdist(as.matrix(scent)), method = "bray"))
```

Mantel test between relatedness and bray curtis similarity of all individuals
```{r}
vegan::mantel(relatedness, scent_bc, method = "spearman", permutation = 1000)
```

Mothers: mantel test between relatedness and bray curtis similarity 
```{r}
vegan::mantel(relatedness[factors$age == 1, factors$age == 1], 
              scent_bc[factors$age == 1, factors$age == 1],
              method = "spearman", permutation = 1000)
```

Pups: mantel test between relatedness and bray curtis similarity 
```{r}
vegan::mantel(relatedness[factors$age == 2, factors$age == 2], 
              scent_bc[factors$age == 2, factors$age == 2],
              method = "spearman", permutation = 1000)
```

***
## Correlation between Heterozygosity and number of compounds in odour profiles

1. Load standardised multilocus heterozygosity (sMLH) based on 41 markers
```{r}
heterozygosity <- as.matrix(read.csv("heterozygosity.csv", row.names=1))
```

2. Average number of compounds per individual
```{r}
num_comp <- as.vector(apply(scent, 1, function(x) length(x[x>0])))
```

3. Correlation between Heterozygosity and number of compounds in mothers
```{r}
het_mum <- heterozygosity[factors$age == 1]
num_comp_mum <- num_comp[factors$age==1]
# summary(glm(het_mum ~ num_comp_mum)))

```

```{r, echo = FALSE, fig.width=5, fig.height=4}
het_df <- as.data.frame(cbind(het_mum, num_comp_mum))
library(ggplot2)
library(grid)
ggplot(het_df, aes(x=het_mum, y=num_comp_mum)) +
        geom_point(colour = "black", size = 2.5) +
        geom_smooth(method="lm",size = 1 ,alpha=0.13, colour="black", fill = "lightblue") +
        theme_minimal(base_size = 18) +
        theme(strip.text.x = element_text(vjust=1,size = 18),
              axis.title.x = element_text(vjust= -2 ,size = 18),
              axis.title.y = element_text(vjust=3,size = 18),
              axis.ticks.x = element_blank(),
              axis.ticks.y = element_blank(),
              plot.margin = (unit(c(.5, .5, 2, 2), "cm"))) +
        scale_x_continuous(breaks=c(seq(from = 0.8, to = 1.20, by = 0.1))) +
        #geom_text(aes(0.85,80, label="(a) r = 0.34, p = 0.027"),size=4) +
        xlab("multilocus heterozygosity") +
        ylab("number of compounds") 
```

4. Correlation between Heterozygosity and number of compounds in pups
```{r}
het_pup <- heterozygosity[factors$age == 2]
num_comp_pup <- num_comp[factors$age==2]
# summary(glm(het_pup ~ num_comp_pup))
```


***resampling plot missing***  
***g2 and plot missing***


