---
title: Fur seal odour encodes colony membership, mother-offspring similarity, relatedness
  and genetic quality
author: Stoffel, M.A., Caspers, B.A., Forcada, J., Giannakara, A., Baier, M.C., Eberhart-Phillips,
  L.J. , Müller, C. & Hoffman, J.I.
output:
  html_document:
    highlight: pygments
  pdf_document: default
---
!!!This file is under developement!!! and will soon contain all major analyses.

This documentation explains the main analyses from our paper and provides the
accompanying code sections as well as figures. The complete code can be found
[in this github repository](https://github.com/mastoffel/scent/tree/master/final_script).
We wrote two small R packages for our analysis. You can directly download them 
from github with the devtools package:

```{r, results="hide", message=FALSE}
library(devtools)
# install_github("mastoffel/minmodelr")
# install_github("mastoffel/g2")
library(minmodelr)
library(g2)
```

See ?MinMod, ?DelTestVar, ?g2 for further information on the functions. We will
use them thoughout this documentation.


***
### Loading data, standardisation and transformation

Loading the raw scent data (aligned by algorithm) and a factor data frame containing identities
for colony membership (colony), mother-offspring pairs (family) and mothers
and pups, respectively (age)

```{r}
scent_raw <- as.data.frame(t(read.csv(".\\files\\scent_raw.csv", row.names = 1)))
factors <- read.csv(".\\files\\factors.csv",row.names=1) 
head(factors)
```

Standardising observations by total, such that within every observation compounds
add up to 100 % (Thus averaging out absolute concentration differences between samples)

```{r}
scent_stand <- as.data.frame(t(apply(scent_raw, 1, function(x) (x/sum(x)) * 100)))
```

Log(x+1) transformation of the standardised scent data.
```{r}
scent <- log(scent_stand + 1)
```

The scent matrix contains 82 observations and 213 compounds (retention times are column names) in total.
```{r}
dim(scent)
head(scent[1:6])
```

***
### Differences in olfactory fingerprints between the two colonies

```{r, results="hide", message=FALSE}
library(vegan)
library(MASS)
```  
    
**Non-metric multidimensional scaling (nMDS)** visualizes a distance matrix (Bray-Curtis similarity).
The nMDS algorithm aims to place each individual in a 2-dimensional space such that the
between-individual distances are preserved as well as possible. Axis coordinates are 
arbitrary and not shown. The plot is better visualized with colours (see paper) and
is shown here for the purpose of demonstration. Mother-offspring pairs can be
identified by labels (e.g. M14, P14)
```{r, fig.height=5, fig.width=5, warning = FALSE}
scent_mds <- isoMDS(vegdist(scent))
ordiplot(scent_mds, type = "t", ylab = "", xlab = "",axes=FALSE, frame.plot=TRUE)
```


**Analysis of Similarities (ANOSIM)** is a distribution-free test for group differences
based on a Bray-curtis similarity matrix. Here is an example which tests for a
significant difference between the two colonies. ANOSIM is philosphically connected
to nMDS as it also uses only the rank order of dissimilarity values.

*ANOSIM - Dissimilarity between the two colonies.*
```{r, cache=TRUE}
anosim(dat = scent, grouping = factors$colony, 
       distance = "bray", permutations = 1000)
```
* The two colonies are highly dissimilar in their olfactory fingerprints.

**Genetic differentiation between the colonies was tested with Structure (Bayesian structure analysis)**  

***
### Mother-offspring similarity

**ANOSIM** can also be used to test for differences between mother-offspring pairs,
basically asking the question if mother-offspring pairs have their own scent "identity".
For computational reasons, the following examples use 1000 permutations while we 
used 10000 for the article.

**Overall**
```{r, cache=TRUE}
anosim(dat = scent, grouping = factors$family, 
       distance = "bray", permutations = 1000)
```

**Within colony 1 (Special study beach)**
```{r, cache=TRUE}

anosim(dat = scent[factors$colony == 1, ], 
       grouping = factors[factors$colony == 1, ]$family, 
       distance = "bray", permutations = 1000)
```
**Within colony 2 (Freshwater beach)**
```{r, cache=TRUE}
anosim(dat = scent[factors$colony == 2, ], 
       grouping = factors[factors$colony == 2, ]$family, 
       distance = "bray", permutations = 1000)
```
 
* Mother-offspring pairs are highly similar in their olfactory fingerprints,
both overall as well as within each colony.

***
### Olfactory similarity vs. geographic distance on special study beach
  
For the special study beach, exact data of the geographical positions is available. 
If mother-offspring similarity would just be caused by spatial proxomity, one would
assume a correlation between geographic distance and olfactory similarity, which is
not the case.
```{r}
coord  <- read.csv(".\\files\\coordinates_beach1.csv", row.names=1) 
```

Converting coordinates to pairwise euclidian distance matrix
```{r}
dist_mat <- as.matrix(dist(coord, method = "euclidian"))
```

Constructing bray curtis similarity matrix of all individuals from beach 1
```{r}
scent_bc <- as.matrix(vegdist(as.matrix(scent[factors$colony == 1, ])),
                      method = "bray")
```

**Geographic distance vs. olfactory similarity in mothers**
```{r}
geo_mum <- dist_mat[1:20, 1:20]
scent_mum <- scent_bc[1:20, 1:20]
vegan::mantel(geo_mum, scent_mum, method = "spearman")
```

**Geographic distance vs. olfactory similarity in pups**
```{r}
geo_pup <- dist_mat[21:40, 21:40]
scent_pup <- scent_bc[21:40, 21:40]
vegan::mantel(geo_pup, scent_pup, method = "spearman")
```

***
## Relatedness and overall olfactory similarity

Due to the olfactory similarity of mothers and their pups, there is the risk
of pseudoreplication for all following analyses. We thus splitted mothers and
pups and analysed them seperately. As we did not find any relationship between
genetic characteristics and scent profiles in pups, the following analysis will
just show the analysis for mother for simplicity and readability.

1. Loading pairwise relatedness based on 41 markers
```{r}
relatedness <- as.matrix(read.csv(".\\files\\relatedness.csv",row.names=1))
```

2. Computing pairwise bray-curtis similarity. 
```{r}
scent_bc <- 1-(as.matrix(vegdist(as.matrix(scent)), method = "bray"))
```

3. **Mantel test between relatedness and bray curtis similarity in mothers** 
```{r}
vegan::mantel(relatedness[factors$age == 1, factors$age == 1], 
              scent_bc[factors$age == 1, factors$age == 1],
              method = "spearman", permutation = 1000)
```
* No relationship between olfactory similarity (based on the overall fingerprint)
and genetic relatedness.

***
## Correlation between Heterozygosity and number of compounds in odour profiles

1. Loading standardised multilocus heterozygosity (sMLH) based on 41 markers
```{r}
heterozygosity <- as.matrix(read.csv(".\\files\\heterozygosity.csv", row.names=1))
```

2. Average number of compounds per individual
```{r}
num_comp <- as.vector(apply(scent, 1, function(x) length(x[x>0])))
```

3. Correlation between Heterozygosity and number of compounds in mothers
```{r}
het_mum <- heterozygosity[factors$age == 1]
num_comp_mum <- num_comp[factors$age==1]
summary(glm(het_mum ~ num_comp_mum))
```

```{r, echo = FALSE, fig.width=6, fig.height=4}
het_df <- as.data.frame(cbind(het_mum, num_comp_mum))
library(ggplot2)
library(grid)
ggplot(het_df, aes(x=het_mum, y=num_comp_mum)) +
        geom_point(colour = "black", size = 2.5) +
        geom_smooth(method="lm",size = 1 ,alpha=0.13, colour="black", fill = "lightblue") +
        theme_minimal(base_size = 18) +
        theme(strip.text.x = element_text(vjust=1,size = 18),
              axis.title.x = element_text(vjust= -2 ,size = 18),
              axis.title.y = element_text(vjust=3,size = 18),
              axis.ticks.x = element_blank(),
              axis.ticks.y = element_blank(),
              plot.margin = (unit(c(.5, .5, 2, 2), "cm"))) +
        scale_x_continuous(breaks=c(seq(from = 0.8, to = 1.20, by = 0.1))) +
        #geom_text(aes(0.85,80, label="(a) r = 0.34, p = 0.027"),size=4) +
        xlab("multilocus heterozygosity") +
        ylab("number of compounds") 
```

## Effect of the number of genetic markers on the correlation between heterozygosity
and olfactory fingerprint complexity

For this study, we used an unusually high number of genetic markers to reliably
estimate heterozygosity and relatedness. We wanted to see how the correlation with 
number of compounds in a scent profile develops with heterozygosity being estimated from an
increasing number of markers. A constant increase also suggests the correlation
to reflect a global rather than a local effect. 

1. Loading genotype data and showing the first 3 loci.
```{r}
library(Rhh)
genotypes <- read.table(".\\files\\genotypes.txt", row.names = 1)
head(genotypes)[1:6]
```

2. Genetic-marker resampling function. 
* genotypes is the genotype data
* y is a vector containing the number of compounds the olfactory fingerprints
* num_iter specifies how often a given set of markers should be randomly sampled
and correlation with heterozygosity computed
* subset_rows subsets mothers after computing heterozygosity from the entire data
```{r}
resample_loci <- function(genotypes, y, num_iter = 1000, subset_rows = 1:41) {
        y <- y[subset_rows]
        num_loci <- ncol(genotypes)/2
        results <- data.frame(matrix(nrow = num_iter, ncol = num_loci))
        ind <- seq(from = 1, to = ncol(genotypes) - 1, by = 2)
        for (i in seq(from = 1, to = num_loci)) {
                for (k in seq_along(1:num_iter)) {
                        loci_ind <- as.list(sample(ind, i, replace = FALSE))
                        gen_sub <- lapply(loci_ind, function(x) genotypes[x:(x+1)])
                        gen_sub <- do.call("cbind", gen_sub)
                        het <- Rhh::sh(gen_sub)[subset_rows, ]
                        results[k, i] <- cor(het,y)
                }
        }   
        results
        }
```

3. Resampling with just 100 iterations instead of 1000 for computational purpose
```{rm, cache=TRUE}
resample_mums <- resample_loci(genotypes, num_comp_mum, num_iter = 10000, subset_rows = 1:41)
```

4. Summing up the results and get mean and se for each marker number
```{r}
sum_results <- function(resampling_output) {
        mean.cor <- apply(resampling_output,2,mean, na.rm=T)
        sd.cor <- apply(resampling_output,2,sd, na.rm=T)
        se.cor <- sd.cor/(sqrt(sd.cor))
        sum_results <- data.frame(locnum = 1:ncol(resampling_output), 
                                  cormean = mean.cor, corsd = sd.cor, corse = se.cor)
}
results_mums <- sum_results(resample_mums)
```

5. Plotting results
```{r, echo=FALSE,  fig.width=6, fig.height=4}
ggplot(results_mums, aes(x = locnum, y = cormean)) +
        geom_errorbar(aes(ymin = cormean-corse, ymax = cormean+corse),
                      width=0.8, alpha=0.7, size = 0.8) +
        geom_point(size = 3, shape = 16) +
        geom_line(size = 0.8) +
        theme_minimal(base_size = 16) +
        theme(axis.title.x = element_text(vjust= -2 ,size = 16),
              axis.title.y = element_text(vjust=3,size = 16),
              axis.ticks.x = element_blank(),
              axis.ticks.y = element_blank(),
              plot.margin = (unit(c(.5, .5, 2, 2), "cm"))) +
        #geom_hline(yintercept=0.305) +
        ylab(expression(r [mean+-se])) +
        xlab("number of loci") +
        ggtitle("Correlation between olfactory complexity and heterozygosity \nestimated by for an increasing number of loci")
```        

## Estimation of second-order heterozygosity disequilibrium with increasing number
of genetic markers




***resampling plot missing***  
***g2 and plot missing***


